---
output: 
  pdf_document:
    toc: true
    number_sections: true
    citation_package: natbib
    highlight: tango
    keep_tex: false
    fig_caption: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "Session 2: SWAT simulation, calibration, and sensitivity analysis in R "
author:
- name: "Christoph Schürz"
  affiliation: "christoph.schuerz@boku.ac.at"
abstract: "The goal of this second session is that you get familiar with the `SWATplusR` package. After the package installation we will load a demo SWAT project and explore the functionality of the `SWATplusR` package. The essential part of this session will be to learn how to utilize the package functionality and combine it with other `R` packages. In a small case study we will execute the SWAT model with different model parametrizations, extract simulation results, evaluate and visualize the simulation results (using the packages `HydroGOF` and the `tidyverse`), perform a parameter sensitivity analysis (employing packages such as `sensitivity` or `fast`), and perform a first model calibration."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# bibliography: ~/Dropbox/master.bib
biblio-style: apsr
---

\newpage
# Package installation
## Package dependencies 
Below you find a list of packages that are required to install `SWATplusR`. Among these packages  you find tools that can be useful in your daily routine when analyzing data in `R` (e.g. most of the packages included in the `tidyverse`). Please install all the packages by executing the following lines of code.

```{r, eval=FALSE}
install.packages(c("doSNOW", "foreach", "RSQLite", "tidyverse"))
```

## SWATplusR
You can install the `SWATplusR` package from my *github* repository (set to public for the course). To install from repositories with ease you can use the package `devtools`
```{r, eval=FALSE}
install.packages("devtools")

# use the function install_github from the devtools package to install
devtools::install_github("chrisschuerz/SWATplusR")
```

## Additional packages
To perform our analyses we require further functionality provided by other `R` packages. Here I provide a list of packages you have to install. I briefly outline what each of these packages does for you.

### `lhs`
This package provides different methods to draw Latin Hypercube Samples. We will sample the SWAT model parameters in the calibration example using latin hypercube sampling.
```{r, eval=FALSE}
install.packages("lhs")
```

### `hydroGOF`
This package provides a comprehensive library of objective criteria used in hydrology (e.g. NSE, KGE, pbias, etc.) to evaluate time series of simulated discharge etc.
```{r, eval= FALSE}
install.packages("hydroGOF")
```

### `sensitivity`
`sensitivity` provides a large variety of methods to perform Global Sensitivity Analysis (GSA, e.g. Sobol or Delsa). 
```{r, eval=FALSE}
install.packages("sensitivity")
```

### fast
The Fourier Amplitude Sensitivity Test (FAST) is a method to perform GSA with few model evaulations. This package impelements this method in `R`.
```{r, eval=FALSE}
install.packages("fast")
```

\newpage

# Load required R packages
```{r, message=F}
library(SWATplusR)
library(tidyverse)
library(lhs)
library(fast)
library(sensitivity)
library(hydroGOF)
```


# Loading the SWAT demo

## Load the project folder
The `SWATplusR` package provides very simple model setups of a head watershed of the Little River Experimental Watershed (LREW). Model setups can be retrievew for SWAT2012 and for SWAT+. The goal is to provide the demos for all operating systems in the future. At the moment the SWAT2012 demo is available for Windows and Linux and the SWAT+ demo is only available for Windows. For this workshop this means that you can work with SWAT+ if you use Windows as your operating system. Linux users have to work with SWAT2012 at the moment. The provided functionality is however very similar for both models.
```{r,eval=FALSE}
# The path where the SWAT demo project will be written
demo_path <- "C:"
# The SWAT version you want to use
swat_version <- "plus" #or "2012" on Linux
# The function writes the demo folder to the defined path and returns the final
# path of the project folder in R
proj_path <- load_demo(dataset = "project", 
                       swat_version = swat_version, 
                       path = demo_path)
```

## Load observation data
The demo contains observation time series data for the main outlet of the demo catchment. Load the table with the observation data and assign it to a variable in your *R* work space.
```{r}
q_obs <- load_demo(dataset = "observation")
```

## First SWAT simulations
After loading the demo project you can already perform your first SWAT simulation. To run SWAT from *R* you can simply use the functions `run_swat2012()` to run a SWAT2012 project or `run_swatplus()` to run a SWAT+ project. We will go though all the parameters these functions provide. The minimum requirement to run simulations in a project is to provide the path to the project and what simulation output the function should return to *R*. Below is a mimimum example to simulate the discharge at the main outlet.

```{r, eval=F}
q_out <- run_swatplus(proj_path, 
                      output = define_output(file = "channel",
                                             variable = "flo_out",
                                             unit = 3))
```

The function returns the simulation at the catchment outlet as a table with a date column and a column with the simulated discharge.

# Changing parameters in a simulation
## A single parameter set
To modify the model parameters in a simulation you simply provide a vector with the parameter values to the `run_swat()` function. A single parameter set can be provided as a vector. To provide an entire set of parameter combinations I recommend to provide them as a tibble. 

For the provided parameter set the naming is very essential and has to follow some rules in order to be interpreted by the function in the correct way. A parameter name can consist of several parts. Some are required, some are optional. The minimum requirement for a parameter name is the actual name in the model an the type of change. Here is an example
```{r}
par_name <- "cn2|change = abschg"
```
The sytax means that we modify the parameter CN2 and we change it by adding an absolute value. If you want to assign an individual name to the parameter in oyur *R* project you can do this as follows:
```{r}
par_name <- "my_name::cn2|change = abschg"
```

Below we define a single parameter set where we reduce all CN2 values by 5 and set the alpha value to 0.5.

```{r}
par_single <- c("cn2.hru|change = abschg" = -5,
                "alpha.gw|change = absval" = 0.5)
```

With this parameter set we can again run our model
```{r, eval=F}
q_out <- run_swatplus(proj_path, 
                      output = define_output(file = "channel",
                                             variable = "flo_out",
                                             unit = 3),
                      parameter = par_single)
```

## Several parameter combinations
The definition of multiple parameter combinations works in a similar way. Now we have to define the parameters with a tibble.
```{r}
par_set <- tibble("cn2.hru|change = abschg" = runif(8,-15,10),
                  "alpha.gw|change = absval" = runif(8, 0, 1))
```
To run SWAT simulations with many parameter sets the `SWATplusR` package provides the option to run the simulations in parallel. To perform parallel simulations you simply define the number of threads as demonstrated below.

```{r, eval=F}
q_out <- run_swatplus(proj_path, 
                      output = define_output(file = "channel",
                                             variable = "flo_out",
                                             unit = 3),
                      parameter = par_set,
                      n_thread = 4)
```
Now instead of only providing a table with the simulations, the function returns by default a list that stores the parameter set that was used, a table that shows details of the parameter definition and the simulation results. As more than one simulation was performed, the columns for the simulation results are now named run_1 to run_8.

