---
output: 
  pdf_document:
    toc: true
    number_sections: true
    citation_package: natbib
    highlight: tango
    keep_tex: false
    fig_caption: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "Session 2: SWAT simulation, calibration, and sensitivity analysis in R "
author:
- name: "Christoph Schürz"
  affiliation: "christoph.schuerz@boku.ac.at"
abstract: "The goal of this second session is that you get familiar with the `SWATplusR` package. After the package installation we will load a demo SWAT project and explore the functionality of the `SWATplusR` package. The essential part of this session will be to learn how to utilize the package functionality and combine it with other `R` packages. In a small case study we will execute the SWAT model with different model parametrizations, extract simulation results, evaluate and visualize the simulation results (using the packages `HydroGOF` and the `tidyverse`), perform a parameter sensitivity analysis (employing packages such as `sensitivity` or `fast`), and perform a first model calibration."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# bibliography: ~/Dropbox/master.bib
biblio-style: apsr
---

\newpage
# Package installation
## Package dependencies 
Below you find a list of packages that are required to install `SWATplusR`. Among these packages  you find tools that can be useful in your daily routine when analyzing data in `R` (e.g. most of the packages included in the `tidyverse`). Please install all the packages by executing the following lines of code.

```{r, eval=FALSE}
install.packages(c("doSNOW", "foreach", "RSQLite", "tidyverse"))
```

## SWATplusR
You can install the `SWATplusR` package from my *github* repository (set to public for the course). To install from repositories with ease you can use the package `devtools`
```{r, eval=FALSE}
install.packages("devtools")

# use the function install_github from the devtools package to install
devtools::install_github("chrisschuerz/SWATplusR")
```

## Additional packages
To perform our analyses we require further functionality provided by other `R` packages. Here I provide a list of packages you have to install. I briefly outline what each of these packages does for you.

### `lhs`
This package provides different methods to draw Latin Hypercube Samples. We will sample the SWAT model parameters in the calibration example using latin hypercube sampling.
```{r, eval=FALSE}
install.packages("lhs")
```

### `hydroGOF`
This package provides a comprehensive library of objective criteria used in hydrology (e.g. NSE, KGE, pbias, etc.) to evaluate time series of simulated discharge etc.
```{r, eval= FALSE}
install.packages("hydroGOF")
```

### `sensitivity`
`sensitivity` provides a large variety of methods to perform Global Sensitivity Analysis (GSA, e.g. Sobol or Delsa). 
```{r, eval=FALSE}
install.packages("sensitivity")
```

### fast
The Fourier Amplitude Sensitivity Test (FAST) is a method to perform GSA with few model evaulations. This package impelements this method in `R`.
```{r, eval=FALSE}
install.packages("fast")
```

\newpage

# Load required R packages
```{r, message=F}
library(SWATplusR)
library(tidyverse)
library(lhs)
library(fast)
library(sensitivity)
library(hydroGOF)
```


# Loading the SWAT demo project
The `SWATplusR` package provides very simple model setups of a head watershed of the Little River Experimental Watershed (LREW). Model setups can be retrievew for SWAT2012 and for SWAT+. The goal is to provide the demos for all operating systems in the future. At the moment the SWAT2012 demo is available for Windows and Linux and the SWAT+ demo is only available for Windows. For this workshop this means that you can work with SWAT+ if you use Windows as your operating system. Linux users have to work with SWAT2012 at the moment. The provided functionality is however very similar for both models.
```{r,eval=FALSE}
# The path where the SWAT demo project will be written
demo_path <- "replace:/with/your/path"
# The SWAT version you want to use
swat_version <- "plus" #or "2012" on Linux
# The function writes the demo folder to the defined path and returns the final
# path of the project folder in R
proj_path <- load_demo(dataset = "project", 
                       swat_version = swat_version, 
                       path = demo_path)
```




